Predicting trends in the quality of state-of-the-art neural networks without access to training or testing data

### 2020-07-02 submission/abstract_file.tex
full pass in light of kdd reviewer comments

In many practical applications, one works with deep neural network (DNN) models trained by someone else.
For such \emph{pretrained models}, one typically does not have access to training data or test data.
Moreover, one does not know many details about the model, such as the specifics of the training data, the loss function, the hyperparameter values, etc.
Given one or many pretrained models, can one say anything about the expected performance or quality of the models?
Here, we present and evaluate empirical quality metrics for pretrained DNN models at scale.
%%Our metrics are drawn from traditional statistical learning theory (e.g., norm-based capacity control metrics) as well as Heavy-Tailed Random Matrix Theory (HT-RMT), as used in the recently-developed Theory of Heavy-Tailed Self Regularization (e.g., fitted power law metrics used to characterize the degree of strong correlations in trained models).
%%The most promising are metrics drawn from traditional statistical learning theory (e.g., norm-based capacity control metrics) as well as metrics (e.g., fitted power law metrics used to characterize the degree of strong correlations in a system) derived from the recently-developed Theory of Heavy-Tailed Self Regularization (HT-SR).
%The most promising are norm-based metrics
% (such as those used to provide capacity control in traditional statistical learning theory) 
%and metrics based on fitting eigenvalue distributions to truncated power law (PL) distributions 
%(which derive from statistical mechanics, in particular the recently-developed Theory of Heavy-Tailed Self Regularization, and which characterize the strength of correlations in a DNN). 
%
Using the open-source \emph{WeightWatcher} tool, we analyze hundreds of publicly-available pretrained models, including older and current state-of-the-art models in computer vision (CV) and natural language processing (NLP).
We examine both familiar norm-based capacity control metrics (Frobenius and Spectral norms) as well as newer Power Law (PL) based metrics (including fitted PL exponents, $\alpha$, and the Weighted Alpha metric, $\hat{\alpha}$), from the recently-developed Theory of Heavy-Tailed Self Regularization (HT-SR).
We also introduce the $\alpha$-Shatten Norm metric.
We find that norm-based metrics correlate well with reported test accuracies for well-trained models across nearly all CV architecture series.
On the other hand, we find that norm-based metrics can not distinguish ``good-versus-bad'' models---which, arguably is the point of needing quality metrics.  
Indeed, they may give spurious results.
We also find that PL-based metrics do much better---quantitatively better at discriminating among a series of ``good-better-best'' models, and qualitatively better at discriminating ``good-versus-bad'' models.
PL-based metrics can also be used to characterize fine-scale properties of these models, and we introduce the layer-wise \emph{Correlation Flow} as new quality assesment.
We show how poorly-trained (and/or poorly fine-tuned) models may exhbit both \emph{Scale Collapse} and unsually large PL exponents, $\alpha\gg 6$, in particular for recent NLP models.
Our techniques, as implemented in the \emph{WeightWatcher} tool, can be used to identify when a pretrained DNN has problems that can not be detected simply by examining training/test accuracies.

>>>>

In many practical applications, one works with deep neural network (DNN) models trained by someone else.
For such \emph{pretrained models}, one often does not have access to training data or test data.
Moreover, one may not know many details about the model, such as the specifics of the training data, the loss function, the hyperparameter values, etc.
Given one or many pretrained models, can one say anything about the expected performance or quality of the models?
Here, we provide a detailed empirical meta-analysis to address this question, evaluating quality metrics for pretrained DNN models at scale.
%%Our metrics are drawn from traditional statistical learning theory (e.g., norm-based capacity control metrics) as well as Heavy-Tailed Random Matrix Theory (HT-RMT), as used in the recently-developed Theory of Heavy-Tailed Self Regularization (e.g., fitted power law metrics used to characterize the degree of strong correlations in trained models).
%%The most promising are metrics drawn from traditional statistical learning theory (e.g., norm-based capacity control metrics) as well as metrics (e.g., fitted power law metrics used to characterize the degree of strong correlations in a system) derived from the recently-developed Theory of Heavy-Tailed Self Regularization (HT-SR).
%The most promising are norm-based metrics
% (such as those used to provide capacity control in traditional statistical learning theory) 
%and metrics based on fitting eigenvalue distributions to truncated power law (PL) distributions 
%(which derive from statistical mechanics, in particular the recently-developed Theory of Heavy-Tailed Self Regularization, and which characterize the strength of correlations in a DNN). 
%
Using the open-source \emph{WeightWatcher} tool to analyze DNN weight matrices, we analyze hundreds of publicly-available pretrained models, including older and current state-of-the-art models in computer vision (CV) and natural language processing (NLP).
We examine both familiar norm-based capacity control metrics (Frobenius and Spectral norms) as well as newer Power Law (PL) based metrics (including fitted PL exponents, $\alpha$, and the Weighted Alpha metric, $\hat{\alpha}$), from the recently-developed Theory of Heavy-Tailed Self Regularization (HT-SR).
We also introduce the $\alpha$-Shatten Norm metric.
We find that norm-based metrics correlate well with reported test accuracies for well-trained models across nearly all CV architecture series.
On the other hand, we find that norm-based metrics can not distinguish well-trained versus poorly-trained models---which, arguably is the point of needing quality metrics.  
%%%Indeed, they may give spurious results.
We also find that PL-based metrics do much better---qualitatively better at discriminating well-trained versus poorly-trained models, and quantitatively better at discriminating among a series of well-trained versus more-well-trained models.
PL-based metrics can also be used to characterize fine-scale properties of these models.
%%%, and we introduce the layer-wise \emph{Correlation Flow} as new quality assessment.
%%%We show how poorly-trained (and/or poorly fine-tuned) models may exhibit both \emph{Scale Collapse} and unusually large PL exponents, $\alpha\gg 6$, in particular for recent NLP models.
Our techniques, as implemented in the \emph{WeightWatcher} tool, can be used to identify when a pretrained DNN has problems that can not be detected simply by examining training/test accuracies.

### 2020-07-02 submission/all_models.tex
full pass, fixed lots of typos, etc.; semifinal, nearly done
full pass in light of kdd reviewer comments

%KDD% \vspace{-1mm}
\section{Comparing Hundreds of CV Models}
\label{sxn:all_cv_models}
%\vspace{-1mm}

In this section, we summarize results from a large-scale analysis of hundreds of CV models, including models developed for image 
classification, segmentation, and a range of related tasks.  
Our aim is to complement the detailed results from Sections~\ref{sxn:cv} and~\ref{sxn:nlp} by providing broader conclusions. % about norm-based and PL-based metrics.
The models we consider have been pretrained on nine datasets.   %, shown in Table~\ref{table:datasets}.
%including ImageNet-1K, and CIFAR-10, CIFAR-100, Street View House Numbers (SVHN), Caltech-UCSD Birds-200-2011 (CUB-200-2011), Pascal VOC2012, ADE20K, Cityscapes, and Common Objects in Context (COCO). 
%%The pretrained models and their accuracy metrics are summarized in the osmr github~\cite{charles-add-link}.
We provide full details about how to reproduce these results in Appendix~\ref{sxn:appendix}.

\begin{table}[t]
\small
\begin{center}
\begin{tabular}{|p{1in}|c|c|c|c|}
%KDD% \begin{tabular}{|p{0.75in}|c|c|c|c|}
\hline
%    & Frobenius Norm & Spectral Norm & Weighted Alpha & Alpha-Norm \\
%    & $\langle\log\Vert\mathbf{W}\Vert_{F}\rangle$ & $\langle\log\Vert\mathbf{W}\Vert_{\infty}\rangle$ & $\langle\hat{\alpha}=\alpha\log\lambda_{max}\rangle$ & $\langle\log\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}\rangle$ \\
Series        & $\log\Vert\cdot\Vert_{F}$ & $\log\Vert\cdot\Vert_{\infty}$ & $\hat{\alpha}$ & $\log\Vert\cdot\Vert^{\alpha}_{\alpha}$ \\
\hline
$R^{2}$ (mean) & 0.63 &  0.55 & \textbf{0.64} & \textbf{0.64} \\
$R^{2}$ (std)  & 0.34 &  0.36 & \textbf{0.29} &          0.30 \\
\hline
$MSE$ (mean)   & 4.54 &  9.62 &          3.14 & \textbf{2.92} \\
$MSE$ (std)    & 8.69 & 23.06 &          5.14 & \textbf{5.00} \\
\hline
\end{tabular}
\end{center}
\caption{Comparison of linear regression fits for different average Log Norm and Weighted Alpha metrics across 5 CV datasets, 17 architectures, covering 108 (out of over 400) different pretrained DNNs.  
         We include regressions only for architectures with five or more data points, and which are postively correlated with test error.
         These results can be readily reproduced using the Google Colab notebooks (see Appendix~\ref{sxn:appendix}). 
        }
\label{table:results}
\end{table}

>>>>

%KDD% 
\vspace{-1mm}
\section{Comparing Hundreds of Models}
\label{sxn:all_cv_models}
%\vspace{-1mm}

In this section, we summarize results from a large-scale analysis of hundreds of publicly-available  models, including models developed for image 
classification, segmentation, and a range of related tasks.  
Our aim is to provide a broad analysis that complements the detailed analysis in Sections~\ref{sxn:cv} and~\ref{sxn:nlp}.
There, we considered CV and NLP models with similar structures. 
Here, we provide much broader conclusions based on a much larger set of models with a much more diverse set of structures. % about norm-based and PL-based metrics.
The models we consider have been pretrained on nine datasets.   %, shown in Table~\ref{table:datasets}.
%including ImageNet-1K, and CIFAR-10, CIFAR-100, Street View House Numbers (SVHN), Caltech-UCSD Birds-200-2011 (CUB-200-2011), Pascal VOC2012, ADE20K, Cityscapes, and Common Objects in Context (COCO). 
%%The pretrained models and their accuracy metrics are summarized in the osmr github~\cite{charles-add-link}.
We provide full details about how to reproduce these results in Appendix~\ref{sxn:appendix}.

\begin{table}[t]
\small
\begin{center}
\begin{tabular}{|p{1in}|c|c|c|c|}
%KDD% \begin{tabular}{|p{0.75in}|c|c|c|c|}
\hline
%    & Frobenius Norm & Spectral Norm & Weighted Alpha & Alpha-Norm \\
%    & $\langle\log\Vert\mathbf{W}\Vert_{F}\rangle$ & $\langle\log\Vert\mathbf{W}\Vert_{\infty}\rangle$ & $\langle\hat{\alpha}=\alpha\log\lambda_{max}\rangle$ & $\langle\log\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}\rangle$ \\
Series        & $\log\Vert\cdot\Vert_{F}$ & $\log\Vert\cdot\Vert_{\infty}$ & $\hat{\alpha}$ & $\log\Vert\cdot\Vert^{\alpha}_{\alpha}$ \\
\hline
$R^{2}$ (mean) & 0.63 &  0.55 & \textbf{0.64} & \textbf{0.64} \\
$R^{2}$ (std)  & 0.34 &  0.36 & \textbf{0.29} &          0.30 \\
\hline
$MSE$ (mean)   & 4.54 &  9.62 &          3.14 & \textbf{2.92} \\
$MSE$ (std)    & 8.69 & 23.06 &          5.14 & \textbf{5.00} \\
\hline
\end{tabular}
\end{center}
\caption{Comparison of linear regression fits for different average Log Norm and Weighted Alpha metrics across 5 CV datasets, 17 architectures, covering 108 (out of over 400) different pretrained DNNs.  
         We include regressions only for architectures with five or more data points, and which are positively correlated with test error.
         These results can be readily reproduced using the Google Colab notebooks (see Appendix~\ref{sxn:appendix}). 
        }
\label{table:results}
\end{table}

### 2020-07-02 submission/background.tex
full pass in light of kdd reviewer comments

Most theory for DNNs is applied to small toy models and assumes access to data.
There is very little work asking how to predict, in a theoretically-principled manner, the quality of large-scale state-of-the-art DNNs, and how to do so without access to training data or testing data or details of the training protocol, etc.
Our 
approach  %%  work 
is, however, 
%% loosely 
related to 
two  %%  several 
other lines of work.

@@

Statistical mechanics ideas have long had influence on DNN theory and practice~\cite{EB01_BOOK, MM17_TR, BKPx20}; and 
our best-performing metrics (those using fitted PL exponents) are based on statistical mechanics~\cite{MM17_TR, MM18_TR, MM19_HTSR_ICML, MM19_KDD, MM20_SDM}, in particular the recently-developed \emph{Theory of Heavy Tailed Self Regularization (HT-SR)}~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}.  
We emphasize that the way in which we (and HT-SR Theory) \emph{use} statistical mechanics theory is quite different than the way it is more commonly formulated.
Several very good overviews of the more common approach are available~\cite{EB01_BOOK, BKPx20}.
We \emph{use} statistical mechanics in a broader sense, drawing upon techniques from quantitative finance and random matrix theory.  % (RMT).
Thus, much more relevant for our methodological approach is older work of Bouchaud, Potters, Sornette, and coworkers~\cite{BouchaudPotters03, SornetteBook, BP11, bun2017} on the statistical mechanics of heavy tailed and strongly correlated systems.
% Probably best not to cite at this point %\charles{, and current work (in progress)\cite{blog}}

@@

There is also a large body of work on using norm-based metrics to bound generalization error~\cite{NTS15, BFT17_TR, LMBx18_TR}.
In this area, theoretical work aims to prove generalization bounds, and applied work uses these norms to construct regularizers to improve training.
While we do find that norms provide relatively good quality metrics, at least for distinguishing good-better-best among well-trained models, we are not interested in proving generalization bounds or developing new regularizers.

>>>>

Most theory for DNNs is applied to small toy models and assumes access to data.
There is very little work asking how to predict, in a theoretically-principled manner, the quality of large-scale state-of-the-art DNNs, and how to do so without access to training data or testing data or details of the training protocol, etc.
%Our 
%approach  %%  work 
%is, however, 
%%% loosely 
%related to 
%two  %%  several 
%other lines of work.
Similarly, there is very little work on performing weight matrix analysis.
Here are several other lines of work most related to our approach.

@@

Statistical mechanics ideas have long had influence on DNN theory and practice~\cite{EB01_BOOK, MM17_TR, BKPx20}; and 
our best-performing metrics (those using fitted PL exponents) are based on statistical mechanics~\cite{MM17_TR, MM18_TR, MM19_HTSR_ICML, MM19_KDD, MM20_SDM}, in particular the recently-developed \emph{Theory of Heavy Tailed Self Regularization (HT-SR)}~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}.  
See Appendix~\ref{sxn:theory-review-appendix} for a brief overview of this theory.
We emphasize that the way in which we (and HT-SR Theory) \emph{use} statistical mechanics theory is quite different than the way it is more commonly formulated.
Several very good overviews of the more common approach are available~\cite{EB01_BOOK, BKPx20}.
We \emph{use} statistical mechanics in a broader sense, drawing upon techniques from quantitative finance and random matrix theory.  % (RMT).
Thus, much more relevant for our methodological approach is older work of Bouchaud, Potters, Sornette, and coworkers~\cite{BouchaudPotters03, SornetteBook, BP11, bun2017} on the statistical mechanics of heavy tailed and strongly correlated systems.
% Probably best not to cite at this point %\charles{, and current work (in progress)\cite{blog}}

@@

There is also a large body of work on using norm-based metrics to bound generalization error~\cite{NTS15, BFT17_TR, LMBx18_TR}.
In this area, theoretical work aims to prove generalization bounds, and applied work uses these norms to construct regularizers to improve training.
While we do find that norms provide relatively good quality metrics, at least for distinguishing among well-trained versus more-well-trained models, we are not interested in proving generalization bounds or developing new regularizers.

\paragraph{Weight matrix analysis.}
To our knowledge, our work, starting in 2018 with the \emph{WeightWatcher} tool~\cite{weightwatcher_package}, is the first to perform a detailed analysis of the weight matrices of DNN models~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}.
Subsequent to the release of the initial version of this paper, we became aware of two other works that were posted to the arXiv in 2020 within weeks of the initial arXiv version of this paper~\cite{EJRUY20_TR,UKGBT20_TR}.
Both of these papers validate our basic result that one can gain substantial insight into model quality by examining weight matrices without access to any training or testing data.
However, both of these papers consider smaller models drawn from a much narrower range of applications than we consider, and previous results in HT-SR Theory suggest that many of the insights from these smaller models may not extend to the state-of-the-art CV and NLP models we consider.

### 2020-07-02 submission/conclusion.tex
full pass in light of kdd reviewer comments

We have developed (based on strong theory) and evaluated (on a large corpus of publicly-available pretrained models from CV and NLP) methods to predict trends in the quality of state-of-the-art neural networks---without access to training or testing data.
Prior to our work, it was not obvious that norm-based metrics would perform well to predict trends in quality \emph{across} models (as they are usually used \emph{within} a given model or parameterized model class, e.g., to bound generalization error or to construct regularizers).
Our results are the first to demonstrate that they can be used for this important practical problem.
That PL-based metrics perform better (than norm-based metrics) should not be surprising---at least to those familiar with the statistical mechanics of heavy tailed and strongly correlated systems~\cite{BouchaudPotters03, SornetteBook, BP11, bun2017} (since our use of PL exponents is designed to capture the idea that well-trained models capture correlations over many size scales in the data).
Again, though, our results are the first to demonstrate this.
It is also gratifying that our approach can be used to provide fine-scale insight (such as rationalizing the flow of correlations or the collapse of size scale) throughout a network.

>>>>

We have developed (based on strong theory) and evaluated (on a large corpus of publicly-available pretrained models from CV and NLP) methods to predict trends in the quality of state-of-the-art neural networks---without access to training or testing data.
Our main methodology involved performing a weight matrix meta-analysis, using the publicly-available \emph{WeightWatcher} tool~\cite{weightwatcher_package}, and informed by the recently-developed HT-SR~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}.
Prior to our work, it was not even obvious that norm-based metrics would perform well to predict trends in quality \emph{across} models (as they are usually used \emph{within} a given model or parameterized model class, e.g., to bound generalization error or to construct regularizers).
Importantly, our results are the first to demonstrate that they can be used for this important practical problem.

Our results also demonstrate that PL-based metrics perform better (than norm-based metrics). 
This should \emph{not} be surprising---at least to those familiar with the statistical mechanics of heavy tailed and strongly correlated systems~\cite{BouchaudPotters03, SornetteBook, BP11, bun2017}---since our use of PL exponents is designed to capture the idea that well-trained models capture information correlations over many size scales in the data.
Again, though, our results are the first to demonstrate this.

It is also gratifying that our approach can be used to provide fine-scale insight (such as rationalizing the flow of correlations or the collapse of size scale) throughout a network. 
Clearly, in light of their role in our results on both CV and NLP models, both \emph{Correlation Flow} and \emph{Scale Collapse} are notions that future work should aim to understand better, both for improved training as well as for improved diagnostics on pretrained~models.
